---
title: "calculate_rates"
author: "Yang Shen"
date: "28-1-2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Updates rates
* mutualism_pars <- list(lac_animal, mu_par, K_par, gam_animal, laa_par, qgain, qloss, lambda1) 
* pars <- c(lac, mu, K, gam, laa), all initial parameters for only one kind of species, here I set for plant species.    
* lac_animal: the initial cladogenesis rates for animal species without mutualism    
* mu_par <- c(mu_A0, mu_P1, mu_A1), in which    
mu_A0: the initial extinction rate of animal species without mutualistic partners  
mu_P1: a coefficient mediate mutualism for plant species extinction rates  
mu_A1: a coefficient mediate mutualism for animal species extinction rates  
* K_par <- c(K_A0, K_P1, K_A1), in which  
K_A0: the initial carrying capacity of animal species without any mutualists  
K_P1: a coefficient showing the effects from mutualism on plant species  
K_A1: a coefficient showing the effects from mutualism on animal species  
* gam_animal: the initial immigration rate for animal species without mutualism  
* laa_par <- c(laa_A0,laa_P1, laa_A1), in which  
laa_A0: the initial anagenesis rates for animal species without mutualism  
laa_P1: a coefficient mediate the effects from mutualism on plant species   
laa_A1: a coefficient mediate the effects from mutualism on animal species  
* qgain: a probability that a pair of plant and animal species could get a link
* qloss: a probability that a pair of plant and animal species could loss a link
* lambda1: a coefficient that mediate the effects from mutualism on cospeciation rates
* M0: the initial interaction matrix on the mainland
* Mt: the interaction matrix on the island at time t
* p_status: a vector showing whether plant species is on the island, 1 means presence, 0 is absence.
* a_status: a vector showing whether animal species is on the island, 1 means presence, 0 is absence.

```{r updates all rates}

update_rates_mutual <- function(gam0_par,
                                lac0_par,
                                K_par,
                                mu_par,
                                laa_par,
                                lambda1,
                                qgain,
                                qloss,
                                M0,
                                Mt,
                                p_status,
                                a_status,
                                island_spec_plant,
                                island_spec_animal) {
  
  part_compe_list <- get_part_compe(M0=M0,
                                    Mt=Mt,
                                    p_status= p_status,
                                    a_status= a_status)
  
  NK_list <- get_NK(K_par=K_par,
                    M0=M0,
                    Mt=Mt,
                    p_status= p_status,
                    a_status= a_status)
  
  expand_matrix_list <- get_expand_matrix(Mt = Mt,
                                          p_status = p_status,
                                          a_status = a_status)
  
  #immigration rate
  immig_rate <- get_immig_rate(
    gam0_par = gam0_par,
    K_par = K_par,
    Mt = Mt,
    M0 = M0,
    p_status = p_status,
    a_status = a_status)
  #testit::assert(is.list(immig_clado_list)) 
  
  #extinction rate
  ext_rate <- get_ext_rate(
    mu_par = mu_par,
    Mt = Mt,
    p_status = p_status,
    a_status = a_status)
  #testit::assert(is.list(ext_rate))
  
  #anagenetic rate
  ana_rate <- get_ana_rate(
    laa_par = laa_par,
    M0 = M0,
    Mt = Mt,
    p_status = p_status,
    a_status = a_status,
    island_spec_plant = island_spec_plant,
    island_spec_animal = island_spec_animal)
  #testit::assert(is.list(ana_rate))
  
  #cladogenetic rate
  clado_rate <- get_clado_rate(
    lac0_par = lac0_par,
    K_par = K_par,
    Mt = Mt,
    M0 =M0,
    p_status = p_status,
    a_status = a_status)
  
  #cospeciation rate
  cospec_rate <- get_cospec_rate(
    lambda1 = lambda1,
    K_par = K_par,
    M0 = M0,
    Mt = Mt,
    p_status = p_status,
    a_status = a_status)
  #testit::assert(is.matrix(cospec_rate))
  
  #gain rate
  gain_rate <- get_gain_rate(
    Mt = Mt,
    p_status = p_status,
    a_status = a_status,
    qgain = qgain)
  #testit::assert(is.matrix(gain_rate))
  
  #loss rate
  loss_rate <- get_loss_rate(
    Mt = Mt,
    p_status = p_status,
    a_status = a_status,
    qloss = qloss)
  #testit::assert(is.matrix(loss_rate))
  
  rates <- list(
    p_immig_rate = immig_rate$plant_immig_rate,
    p_ext_rate = ext_rate$plant_ext_rate,
    p_clado_rate = clado_rate$plant_clado_rate,
    p_ana_rate = ana_rate$plant_ana_rate,
    a_immig_rate = immig_rate$animal_immig_rate,
    a_ext_rate = ext_rate$animal_ext_rate,
    a_clado_rate = clado_rate$animal_clado_rate,
    a_ana_rate = ana_rate$animal_ana_rate,
    cospec_rate = cospec_rate,
    gain_rate = gain_rate,
    loss_rate = loss_rate)
  
  return(rates)
}
```

# Calculates rates according to the equations
## calculate mutualistic partners and competitors for plant and animal species
```{r partners and competitors}
# example:
# Mt <- {set.seed(1);matrix(sample(c(0,1),20,replace = TRUE),ncol=5,nrow=4)}
# p_status<-c(0,1,1,0)
# a_status<-c(1,0,0,0,1)
# part_compe_list <-get_part_compe(Mt, p_status,a_status)
get_part_compe <- function(Mt,
                           p_status,
                           a_status){
  tMt <- t(Mt)
  
  plant_partners <- Mt %*% a_status #the number of partners that each plant species has
  animal_partners <- tMt %*% p_status #the number of partners that each animal species has
  
  plant_compe <- matrix()
  animal_compe <- matrix()
  
  for (x in seq(NROW(Mt))){
    plant_compe[x] <- sum((colSums(tMt[,x]*tMt[,-x]* a_status)>=1) * as.matrix(p_status)[-x,])
  }  # the number of competitors of each plant species
  
  for (x in seq(NCOL(Mt))){ 
    animal_compe[x] <- sum((colSums(Mt[,x] * Mt[,-x]* p_status)>=1) * as.matrix(a_status)[-x,])
  } # the number of competitors of each animal species
  
  part_compe_list <- list(plant_partners = plant_partners,     
                          animal_partners = animal_partners,   
                          plant_compe = plant_compe,   
                          animal_compe = animal_compe) 

  return(part_compe_list)
}
```
## calculate N/K in mutualistic system
```{r get N/K in mutualistic system}
get_NK <- function(K,
                   Mt,
                   p_status,
                   a_status,
                   mutualism_pars){
  part_compe_list <- get_part_compe(Mt = Mt,
                                    p_status = p_status,
                                    a_status = a_status)
  
  indp <- which(part_compe_list[[1]]==0)
  inda <- which(part_compe_list[[2]]==0)
  
  K_par <- mutualism_pars$K_par
  
      plant_NK <- exp(-(sum(p_status)/K+
                           part_compe_list[[3]]/(K_par[2]*part_compe_list[[1]])))
      plant_NK[indp] <- exp(-(sum(p_status)/K)) # N/K for plant species. According to the equation,
                                # If there's no mutualistic partners on the island, N/K should be 
                                # exp(-N/K), otherwise it's exp(-(N/K+N'/(K'*partners))).
 
      animal_NK <- exp(-(sum(a_status)/K_par[1]+
                            part_compe_list[[4]]/(K_par[3]*part_compe_list[[2]]))) 
      animal_NK[inda] <- exp(-(sum(a_status)/K_par[1]))# N/K for animal species
  
  NK_list <- list(plant_NK = plant_NK,
                  animal_NK = animal_NK)
  
  return(NK_list)
}

```
## expand matrix
```{r expand matrix for easying on calculation}
get_expand_matrix <- function(Mt,
                              p_status,
                              a_status){
  expd_p_sta <- matrix(rep(p_status,NCOL(Mt)), ncol = NCOL(Mt))
  expd_a_sta <- t(matrix(rep(a_status,NROW(Mt)),ncol = NROW(Mt)))
  
  expand_matrix_list <- list(expd_p_sta = expd_p_sta,
                             expd_a_sta = expd_a_sta)
  return(expand_matrix_list)
}
```
## calculate immigration rates
```{r immigration rates}
get_immig_rate <- function(gam,
                           K,
                           M0,
                           Mt,
                           p_status,
                           a_status,
                           mutualism_pars){
  
  if (is.null(mutualism_pars)){
    mainland_n <- sum(NROW(M0)+NCOL(M0)) #the number of species on the mainland
    num_spec <- sum(p_status)+sum(a_status) #the number of species on the island
    immig_rate <- max(c( mainland_n * gam * (1 - (num_spec / K)),
                         0), na.rm = TRUE)
    return(immig_rate)
  } else {
    NK_list <- get_NK(K = K,
                      Mt = Mt,
                      p_status = p_status,
                      a_status = a_status,
                      mutualism_pars = mutualism_pars)
    
    gam_animal <- mutualism_pars$gam_animal
    
    plant_immig_rate <- gam * NK_list[[1]][1:NROW(M0)] #NK_list calculates for all                                                             #species, but only those ID from                                                         #mainland can immigrate.                                                            #rows for plant while cols for animal. 
    animal_immig_rate <- gam_animal * NK_list[[2]][1:NCOL(M0)] 
    
    immig_list <- list(plant_immig_rate = plant_immig_rate,
                       animal_immig_rate = animal_immig_rate)
    
    return(immig_list)
  }
}
```
## calculate anagenesis rates
```{r anagenesis rates}
get_ana_rate <- function(laa,
                         num_immigrant,
                         M0,
                         Mt,
                         p_status,
                         a_status,
                         mutualism_pars) {
  
  if(is.null(mutualism_pars)){
    ana_rate <- laa * num_immigrants
    return(ana_rate)
  }else{
    
    laa_par <- mutualism_pars$laa_par
    
    plant_ana_rate = laa + laa_par[2] * (abs(Mt[1:NROW(M0),1:NCOL(M0)]-M0) %*% a_status[1:NCOL(M0)])  #"[1:NROW(M0),1:NCOL(M0)]" keeps only species ID from mainland                          #can happend to anagenesis.                                                                                     
    animal_ana_rate = laa_par[1] + laa_par[3] * (t(abs(Mt[1:NROW(M0),1:NCOL(M0)]-M0)) %*%                                                               p_status[1:NROW(M0)]) 
    
    ana_list <- list(plant_ana_rate = plant_ana_rate,
                     animal_ana_rate = animal_ana_rate)
    return(ana_list)
  }
}

```
## calculate cladogenesis rates
```{r get cladogenesis rates}
get_clado_rate <- function(lac,
                           K,
                           Mt,
                           p_status,
                           a_status,
                           mutualism_pars){
  
  if (is.null(mutualism_pars)){
    num_spec <- sum(p_status)+sum(a_status)
    clado_rate <- max(
      0, lac * num_spec  * (1 - num_spec /K), na.rm = TRUE)
    return(clado_rate)
  }else{
    
    NK_list <- get_NK(K = K,
                      Mt = Mt,
                      p_status = p_status,
                      a_status = a_status,
                      mutualism_pars = mutualism_pars)
    
    lac_animal <- mutualism_pars$lac_animal
    
    plant_clado_rate <- lac * NK_list[[1]] * p_status #"* p_status" assures the                                                           #species is on the island 
    animal_clado_rate <- lac_animal * NK_list[[2]] *a_status
    #only species presented on island have cladogenestic rates
    
    clado_list <- list(plant_clado_rate = plant_clado_rate,
                       animal_clado_rate = animal_clado_rate)
    
    return(clado_list)
  }
}
```
## calculate extinction rates
```{r get extinction rates}
get_ext_rate <- function(mu,
                         Mt,
                         p_status,
                         a_status,
                         mutualism_pars) {
  
  if (is.null(mutualism_pars)){
    num_spec <- sum(p_status)+sum(a_status)
    ext_rate <- max(0, mu * num_spec, na.rm = TRUE)
    return(ext_rate)
  } else {
  part_compe_list <- get_part_compe(Mt=Mt,
                                    p_status= p_status,
                                    a_status= a_status)
  mu_par <- mutualism_pars$mu_par
  
  plant_ext_rate <- as.matrix(pmax(0, mu - mu_par[2] * part_compe_list[[1]])) * p_status
  animal_ext_rate <- as.matrix(pmax(0, mu_par[1] - mu_par[3] * part_compe_list[[2]])) *a_status
  
  ext_list <- list(plant_ext_rate = plant_ext_rate,
                   animal_ext_rate = animal_ext_rate)
  return(ext_list)
  }
}
```
## calculate cospecition rates
```{r get cospeciation rates}
get_cospec_rate <- function(K,
                            Mt,
                            p_status,
                            a_status,
                            mutualism_pars){
  
  NK_list <- get_NK(K = K,
                    Mt = Mt,
                    p_status = p_status,
                    a_status = a_status,
                    mutualism_pars = mutualism_pars)
  
  expand_matrix_list <- get_expand_matrix(Mt = Mt,
                                          p_status = p_status,
                                          a_status = a_status)
  lambda1 <- mutualism_pars$lambda1
  cospec_rate <- lambda1 * Mt * expand_matrix_list[[1]] * expand_matrix_list[[2]] *
    matrix(rep(NK_list[[3]],NCOL(Mt)), ncol = NCOL(Mt)) *
    t(matrix(rep(NK_list[[4]],NROW(Mt)),ncol = NROW(Mt)))
  
  return(cospec_rate)
}

```
## calculate gain rates
```{r get gain rates}
get_gain_rate <- function(Mt,
                          p_status,
                          a_status,
                          mutualism_pars){
  
  expand_matrix_list <- get_expand_matrix(Mt = Mt,
                                          p_status = p_status,
                                          a_status = a_status)
  qgain <- mutualism_pars$qgain
  gain_rate <-  qgain * (1-Mt) * expand_matrix_list[[1]] * expand_matrix_list[[2]]
  
  return(gain_rate)
}
```
## calculate loss rates
```{r get loss rates}
get_loss_rate <- function(Mt,
                        p_status,
                        a_status,
                        mutualism_pars){
  
  expand_matrix_list <- get_expand_matrix(Mt = Mt,
                                          p_status = p_status,
                                          a_status = a_status)
  qloss <- mutualism_pars$qloss
  
  loss_rate <- qloss * Mt * 
    (expand_matrix_list[[1]]*expand_matrix_list[[2]] + 
       expand_matrix_list[[2]]*(1-expand_matrix_list[[1]]) + 
       expand_matrix_list[[1]]*(1-expand_matrix_list[[2]]))
  
  return(loss_rate)
}
```

```{r cars}

```















